Behind Bars: Insights into the Wellbeing of U.S. Prisoners

### 2024-02-16 paper/paper.qmd

According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside
Federal Bureau of Prisons(BoP) institutions between 2014 and 2021. The staff providing psychiatric services in the prisons
stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary
resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that
time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths [@Levin_2024].
News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics.
This has sparked calls for reform and raised concerns about the efficacy of healthcare in prisons.

@@

As we dig deeper, our focus goes beyond statistics, aiming to amplify the voices and experiences of those directly
affected. This exploration goes beyond the numbers, delving into the personal dimensions of the prison experience and
laying the groundwork for a comprehensive analysis that seeks to drive change in a system facing significant challenges.

@@

- x1 (The race of the prisoner)
- Cause (the cause of death of the prisoner)
- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)

@@

- x1 (The gender of the prisoner)
- Cause (the cause of death of the prisoner)
- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)

@@

- Date (The date of the new case)
- New_cases_per_100k (The number of new cases recorded on the specific date per 100k)
- Group (The group of people the case belonged to)

@@

## Weaknesses and next steps

>>>>

According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside Federal Bureau of Prisons(BoP) institutions between 2014 and 2021. The staff providing psychiatric services in the prisons stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths [@Levin_2024].

News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics. This has sparked calls for reform and raised concerns about the efficacy of healthcare in prisons.

@@

As we dig deeper, our focus goes beyond statistics, aiming to uncover trends of causes of prisoner death rates in the United States and examine these death rates both from a comprehensive standpoint and delve into demographic breakdowns, specifically considering gender and race. We also draw in Covid-19 data about new cases to support our analysis of wellbeing in prison.

The paper is further organised into three sections: Data, Results, and Discussion. In the Data section, I discuss how the datasets to be used for the analysis were obtained, pre-processed and cleaned. I will also explain the variables of interest in the datasets for my analysis. The Results section will then highlight and discuss the trends and associations found during the analysis. Lastly, the Discussion section will talk about some interesting trends found in Results in depth, link it to the real world and also highlight the weaknesses of our analysis.

@@

- x1 (the race of the prisoner)
- Cause (the cause of death of the prisoner)
- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)

@@

- x1 (the gender of the prisoner)
- Cause (the cause of death of the prisoner)
- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)

@@

- Date (the date of the new case)
- New_cases_per_100k (the number of new cases recorded on the specific date per 100k)
- Group (the group of people the case belonged to)

@@

## Weaknesses
While our analysis of prisoner mortality provides valuable insights, it is important to acknowledge certain limitations. The binary representation of gender in the original paper oversimplifies a complex spectrum, neglecting nuances beyond the male-female binary.

Additionally, the absence of data beyond 2016 restricts our ability to capture recent developments, and the lack of detailed information about other races limits a comprehensive understanding of racial disparities.

Also, the datasets’ focus on co-ed adult prisons overlooks the diverse landscape of correctional facilities, including juvenile and gender-specific institutions, which could yield distinct insights.

Furthermore, the dataset gives collective information about the prisons in the US. The inclusion of regional split information could help us uncover regional variations in behavior and pre-existing medical conditions to name a few factors.

Lastly, exploring the intersectionality of race and gender would provide a more nuanced understanding of the complex factors influencing prisoner mortality.

## Next Steps
To take our study one step beyond, we could look for datasets that would fulfill the weaknesses of our datasets and combine that with our existing data. This could strengthen the general investigation that we are trying to make. We could also explore the data of other countries with either similar geography or demographic to compare how each of the prison systems perform under the general well-being category.
